Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-20 paper/paper.qmd
update reference and introduction of paper

---
title: "Examining the Overturn of Roe v. Wade: The Impact of Abortion Bans on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.268 deaths per 1,000 people in states that became abortion-banned, compared to states where abortion remained legal based on Difference in Difference linear modeling."
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

```{r}
analysis_data<-read_parquet(here::here("data", "02-analysis_data", "analysis_data.parquet"))
```

@@

The overturning of Roe v. Wade on June 24, 2022, in the Supreme Court’s Dobbs v. Jackson Women’s Health Organization decision ended nearly five decades of constitutional protection for abortion rights. This ruling returned authority over abortion regulations to individual states, resulting in a patchwork of laws across the country. The shift sparked debates over reproductive rights and public health, underscoring disparities in access to abortion services and their impact on health outcomes in states with restrictive policies.

The banning of abortion not only raises maternal mortality rates—evidenced by a 56% increase in Texas from 2019 to 2022 [@nbcnews]—but also influences infant mortality rates. Restrictive abortion laws can lead to more unintended or high-risk pregnancies, particularly among disadvantaged individuals with limited healthcare access, resulting in higher rates of preterm births, low birth weights, and delivery complications that elevate infant mortality.

Before

The estimand of our analysis is the change in the difference in infant death rates between states that banned abortion and states where abortion remained legal after June 24, 2022. The object of the estimation is to estimate the change in the difference in the infant death rate between the treatment and control groups based on observational data in the United States using the Difference-in-Difference approach. Our aim is to identify changes in infant mortality trends over time, providing a clearer understanding of the potential impacts of abortion bans on public health outcomes.

The structure of this paper is organized as follows: @sec-data 
provides details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions
and specifications of our linear regression framework. In
@sec-result, we present our findings, emphasizing the key predictors
of Harris's support. Finally, @sec-discussion explores the
implications of our results and suggests potential directions for future
research. @sec-a provides external data detail, @sec-b provides
model detail, @sec-c provides an investigation for pollster Siena College's
methodology, and  @sec-d provides an idealized methodology for polling the poll.

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

We conduct our infant death rate observational data analysis using the R programming language
[@citeR]. Our dataset, obtained from CDC Wonder
[@cdc_wonder_2024], based on data as of December 2022, provides
a detailed overview of public opinion in the lead-up to the election.
Adhering to the guidelines presented in @tellingstories, we explore
various factors that influence voter support percentages, including the
timing of the polls, the traits of polling organizations, and regional
differences.

Overview text

@@

### Abortion Legality & After Injunction

@@

Another created variable `after_injunction` indicates whether the date of infants' death falls after the implementation of the abortion restrictions following the overturn of Roe v. Wade in June 2022. It is coded as a binary variable, where 1 represents dates after June 2022, signifying the post-injunction period, and 0 represents dates before June 2022, indicating the pre-injunction period.

@@

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State. As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group)) +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  ) +
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) 
```
@fig-map illustrates the legality of abortion across U.S. states. Based on @reproductiveright, as of November 2024, abortion has been illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin following the June 2022 decision.
```{r}
#| label: fig-rateandstate
#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
#| echo: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() 
```
@fig-rateandstate illustrates trends in infant death rates (per 1,000 live births) over time in states where abortion is illegal versus legal. In states with abortion bans, infant mortality shows a slight increase after June 2022, marked by the dashed line representing the overturn of Roe v. Wade. In contrast, states where abortion remains legal exhibit relatively stable trends. This suggests a potential link between abortion bans and rising infant mortality rates, warranting further investigation.

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>
  group_by(state)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`State`=state)

>>>>

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.31 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

```{r}
#| include: false
#| warning: false
#| message: false
analysis_data<-read_parquet(here::here("data", "02-analysis_data", "analysis_data.parquet"))
```

@@

The overturning of Roe v. Wade on June 24, 2022, through the Supreme Court’s decision in Dobbs v. Jackson Women’s Health Organization, ended nearly five decades of constitutional protection for abortion rights. This landmark ruling shifted authority over abortion regulations to individual states, creating a patchwork of laws nationwide. The decision reignited debates about reproductive rights and public health, highlighting disparities in access to abortion services and their effects on health outcomes, especially in states with restrictive policies.

Banning abortion usually has been linked to a rise in maternal mortality rates, with Texas experiencing a 56% increase from 2019 to 2022 [@nbcnews]. However, restrictive abortion laws can also impact infant mortality, as they often lead to more unintended or high-risk pregnancies among disadvantaged populations with limited access to healthcare. These pregnancies are associated with higher rates of preterm births, low birth weights, and delivery complications, all of which elevate infant mortality rates. Even before Roe v. Wade was overturned, researchers highlighted the impacts of restrictive abortion laws on mortality rates. @Burdick2024 found that states with 11–12 restrictive laws during 2014–2018 had a 16% higher infant mortality rate, particularly in counties with higher proportions of Black populations, inadequate prenatal care, and maternal smoking. Similarly, @Harper2023 showed that restrictive laws are linked to higher maternal, fetal, and infant mortality, states with moderate abortion legislation had significantly lower maternal mortality (25.79 per 100,000 live births) and infant mortality (20.56 per 1,000 live births) compared to restrictive states, underscoring the health benefits of less restrictive policies.

The estimand of our analysis is the change in the difference in infant death rates between states that banned abortion and states where abortion remained legal after June 24, 2022.  The object of the estimation is to estimate the change in the difference in the infant death rate between the treatment and control groups based on observational data in the United States.

Instead of examining the impact of abortion laws over broader time periods like prior research, our study focuses on the effect of abortion bans on infant mortality rates in the United States specifically from 2021 to 2022, with an emphasis on changes following the 2022 Dobbs decision. Using a Difference-in-Differences (DID) approach, we assess the interaction between abortion legality in states and the timing of the injunction implemented to understand its influence on infant health outcomes. The study utilizes observational data from the Centers for Disease Control and Prevention’s WONDER database, capturing infant mortality trends across states with varying levels of abortion restrictiveness. Key predictors include the mother’s state of residence, race, age, and the year of infant death, enabling a nuanced examination of the effects of restrictive abortion laws on public health.

Our Results show that the overturning of Roe v. Wade and the implementation of abortion bans are linked to a rise in infant mortality rates in states where abortion became illegal. Through a Difference-in-Differences (DID) approach, we estimate that states enforcing abortion bans after the injunction saw an increase of 0.31 deaths per 1,000 live births compared to states where abortion remained legal. Furthermore, we find that maternal age and race are key determinants of infant mortality, with older maternal age groups associated with lower mortality rates and Black mothers experiencing significantly higher rates. These results emphasize the public health consequences of restrictive abortion laws and the urgent need for policies that mitigate these disparities.

This study is important as we highlight the public health implications of abortion bans, particularly their role in increasing infant mortality rates. By focusing on the 2022 Dobbs decision, we provide timely evidence on how restrictive abortion policies disproportionately affect vulnerable populations, emphasizing the need for data-driven policies to mitigate these adverse outcomes.

The structure of this paper is organized as follows: @sec-data  provides details on the data sources and variables used in our analysis. @sec-model explains the modeling approach, including the assumptions and specifications of our linear regression framework. In @sec-result, we present our findings, emphasizing the key predictors of Harris's support. Finally, @sec-discussion explores the implications of our results and suggests potential directions for future research. @sec-a provides external data detail, @sec-b provides model detail, @sec-c provides an datasheet, and  @sec-d provides an detailed introduction of surveys, sampling and observational data in our paper.

@@

We conduct our infant death rate observational data analysis using the R programming language
[@citeR]. Our dataset, obtained from CDC Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a detailed overview of the Infant Death Rate across the USA. CDC WONDER (Wide-ranging ONline Data for Epidemiologic Research) is an online system providing public health professionals and the public with access to a wide range of the Centers for Disease Control and Prevention (CDC) epidemiologic data and resources. Following the guidelines outlined in @tellingstories, we analyze the impact of abortion injunctions on infant death rates using the Difference-in-Difference method. This approach utilizes the interaction between the legality of abortion in the mother’s state of residence and the timing of the infant’s death as the predictor variable. Additionally, we include control variables such as infant year of death, the mother’s state of residence, race, and age to account for potential confounding factors.

@@

### Interaction Term: Abortion Legality and After Injunction

@@

Another created variable `after_injunction` indicates whether the date of infants' death falls after the implementation of the abortion injunctions following the overturn of Roe v. Wade in June 2022. It is coded as a binary variable, where 1 represents dates after June 2022, signifying the post-injunction period, and 0 represents dates before June 2022, indicating the pre-injunction period.

To evaluate the impact of abortion injunctions on infant death rates, we utilize the Difference-in-Difference (DID) method. As part of this approach, we construct an interaction term defined as the product of `abortion_illegal` and `after_injunction`. The resulting interaction term equals 1 when abortion is illegal and the time is after the injunction, and 0 otherwise. This term is included in the analysis to evaluate the joint impact of these factors on infant death rates.

@@

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State. As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) 
```
@fig-map illustrates the legality of abortion across U.S. states. Based on @reproductiveright, as of November 2024, abortion has been illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin following the June 2022 decision.
```{r}
#| label: fig-rateandstate
#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
#| echo: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() 
```
@fig-rateandstate illustrates trends in infant death rates (per 1,000 live births) over time in states where abortion is illegal versus legal. In states with abortion bans, infant mortality shows a slight increase after June 2022, marked by the dashed line representing the overturn of Roe v. Wade. In contrast, states where abortion remains legal exhibit relatively stable trends. This suggests a potential link between abortion bans and rising infant mortality rates, warranting further investigation.

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing. States with “Illegal Abortion and After Injunction” (denoted as ‘1’) have a higher average infant death rate compared to states with “Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction” (denoted as ‘0’).
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`'0': Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction, '1': Illegal Abortion and After Injunction`=illegal_after_injunction)

# Display the result as a table
kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
```
@tbl-interact highlights average infant death rates (per 1,000 live births) based on abortion legality and injunction timing. States categorized as ‘1’, where abortion became illegal after the injunction, show a higher average infant death rate of 8.3 compared to 7.9 in ‘0’ states, where abortion remained legal or abortion used to be legal before the injunction. This suggests that restrictive abortion policies and their enforcement may contribute to higher infant mortality, potentially due to increased unintended or high-risk pregnancies, reduced access to maternal healthcare, and exacerbated social or economic disparities.

## Control Variables

@@

```{r}
#| label: fig-choromap
#| fig-cap: Infant Death Rates Across U.S. States (per 1,000 Live Births), December 2022. Infant Mortality Rate is much higher iin Southeastern. Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
#| echo: false

library(ggnewscale)
death<-analysis_data|>group_by(state)|>summarise(death_rate=mean(death_rate))|>mutate(state=tolower(state))
death_full <- us_map|>
   full_join(death, c("region"="state"))

ggplot() +
  # First layer: Fill with death_rate
  geom_polygon(
    data = death_full,
    aes(x = long, y = lat, group = group, fill = death_rate),
    color = "black"
  ) +
  scale_fill_viridis_c(
    name = "Infant Death Rate\n(per 1,000 live births)",
    option = "C",
    na.value = "grey50",
    breaks = c(10, 15, 20,25),  # Custom breaks for the color legend
  ) +
  theme_void() +
  labs(
    caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
  ) 
```
@fig-choromap is a choropleth map visualizes the average infant death rate per 1,000 live births across U.S. states, with data categorized by color intensity. States with higher infant death rates are shown in lighter shades (yellow), while states with lower rates are represented in darker shades (purple). States with missing data are indicated in grey. The map highlights disparities in infant mortality rates, with some states, particularly in the southeastern U.S., exhibiting notably higher rates compared to others. States like Alabama and Mississippi show the highest rates, while states such as California and New York exhibit lower rates.
